Flabby and injective objects in toposes

### 2018-07-06 paper-flabby-objects.tex
typo

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study its basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. Continuing a line of
  research started by Roswitha Harting, we use this notion to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. As an application, we give internal characterizations of
  sheaf cohomology and show that higher direct images can be understood as
  internal cohomology.
\end{abstract}

@@

@@

Both definitions can be generalized to yield notions of flabby objects in
elementary toposes; but for toposes which are not localic, the two resulting
notions will differ, and only the one induced by the alternative definition
can be characterized in the internal language. We therefore adapt in this paper the
alternative one as the official definition.

@@

\begin{prop}\label{prop:flabby-sheaves-objects}
Let~$F$ be a sheaf on a topological space~$X$ (or a locale).
Then~$F$ is flabby as a sheaf if and only if~$F$ is a flabby object of the
sheaf topos~$\Sh(X)$.
\end{prop}

@@

\begin{prop}\label{prop:global-elements}
Let~$X$ be a flabby object in an elementary topos~$\E$. If~$\E$ is localic and
Zorn's lemma is available, then~$X$ possesses a global element (a morphism~$1 \to X$).
\end{prop}

@@

The analogue of Theorem~\ref{thm:injectivity-external-internal} for modules
holds as well, if~$\E$ is assumed to have a natural numbers object. The proof
carries over word for word, only referencing
Proposition~\ref{prop:set-of-extensions-flabby} instead of
Proposition~\ref{prop:set-of-linear-extensions-flabby}.
It seems that Harting was not aware of this, even though she did show that
injectivity of sheaves of modules over topological spaces is a local
notion~\cite{XXX}, as she
(mistakingly) states in~\cite[page~233]{harting:remark} that ``the notions of
injectivity and internal injectivity do not coincide'' for modules.
% XXX "mistakenly"? comparison with Harting's proof (Barr's lemma...)

@@

\begin{scholium}\label{scholium:properties-of-flabby-objects}
For any elementary topos~$\E$:
\begin{enumerate}
\item Any object can be embedded into an internally injective object.
\item (If~$\E$ has a natural numbers object.) The underlying unstructured
object of an internally injective module is internally injective.
\item Any internally injective object is flabby.
% Take care if numbering changes (referenced below).
\item Any object can be embedded into a flabby object.
\item The terminal object is flabby. The product of flabby objects if flabby.
\item Let~$I$ be an internally injective object. Let~$T$ be an arbitrary
object. Then~$[T,I]$ is a flabby object.
\item (If~$\E$ has a natural numbers object.) Let~$I$ be an internally
injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then~$[T,I]_R$, the
subobject of the internal Hom consisting only of the linear maps, is a flabby
object.
\item Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
of~$R$-modules in~$\E$. If~$M'$ and~$M''$ are flabby objects, so is~$M$.
\end{enumerate}
\end{scholium}

@@

The notion of strongly flabby objects is, however, not stable under base change
and therefore cannot be characterized in the internal language. A specific
example is the~$G$-set~$G$ (with the translation action), considered as an
object of the topos~$\PSh(BG)$ of~$G$-sets, where~$G$ is a nontrivial group.
This object is not strongly flabby, since the morphism~$\emptyset \to G$ does
not lift, but its pullback to the slice~$\PSh(BG)/G \simeq \Set$ is (assuming
the law of excluded middle).
\end{rem}

@@

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. It is therefore instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the effective
topos.

@@

We follow the terminology of~\cite{hyland:effective-topos}(XXX). The intuitive
reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
following. Let~$X$ be a flabby object of the effective topos. Then there is a
procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
on~$K$, but this fine point is not important for the further discussion. Let
now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.

@@

This problem does not manifest with objects~$X$ which are not effective sets.
Reflexivity realizers for these objects may not be very informative themselves;
a procedure witnessing flabbiness therefore only has to turn one kind of
non-informative realizers into another kind. The flabby modules featuring in
the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
be not effective sets.

@@

\item There is a way of embedding any module into a flabby module if one is
prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
Inductively construct a collection~$T$ of terms by the following clauses: $0
\in T$ (where~$0$ is a formal symbol);
if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $1t \sim t$, $r(t+s) \sim rt
+ rs$, $(r+r')t \sim rt + r't$ and such
that~$\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
r \underline{x}$, $\underline{0} \sim 0$ and such that~$\varepsilon_{\{t\}} \sim
t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.

\item There appears to be some tension regarding the effective topos:
Proposition~\ref{prop:semienough-flabby-modules} shows that at
least~$\neg\neg$-separated modules in the effective topos always embed into
flabby modules, assuming the law of excluded middle, while
Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective
module is flabby.
\end{enumerate}

>>>>

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study their basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. Continuing a line of
  research started by Roswitha Harting, we use flabby objects to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. As an application, we give internal characterizations of
  sheaf cohomology and show that higher direct images can be understood as
  internal cohomology.
\end{abstract}

@@

% XXX: In Einleitung interne Sprache erinnern.

@@

Both definitions can be generalized to yield notions of flabby objects in
elementary toposes; but for toposes which are not localic, the two resulting
notions will differ, and only the one obtained from the alternative definition
is stable under pullback and
can be characterized in the internal language. We therefore adapt in this paper the
alternative one as the official definition.

@@

\begin{prop}\label{prop:flabby-sheaves-objects}
Let~$F$ be a sheaf on a topological space~$X$ (or a locale).
Then~$F$ is flabby as a sheaf if and only if~$F$ is flabby as an object of the
sheaf topos~$\Sh(X)$.
\end{prop}

@@

\begin{prop}\label{prop:global-elements}
Let~$X$ be a flabby object in an elementary topos~$\E$. If~$\E$ is localic and
Zorn's lemma is available in the metatheory, then~$X$ possesses a global element (a morphism~$1 \to X$).
\end{prop}

@@

The analogue of Theorem~\ref{thm:injectivity-external-internal} for modules
holds as well, if~$\E$ is assumed to have a natural numbers object. The proof
carries over word for word, only referencing
Proposition~\ref{prop:set-of-extensions-flabby} instead of
Proposition~\ref{prop:set-of-linear-extensions-flabby}.
It seems that Harting was not aware of this, even though she did show that
injectivity of sheaves of modules over topological spaces is a local
notion~\cite{XXX}, as she
(mistakenly) states in~\cite[page~233]{harting:remark} that ``the notions of
injectivity and internal injectivity do not coincide'' for modules.
% XXX comparison with Harting's proof (Barr's lemma...)

@@

\begin{scholium}\label{scholium:properties-of-flabby-objects}
For any elementary topos~$\E$:
\begin{enumerate}
\item Any object can be embedded into an internally injective object.
\item (If~$\E$ has a natural numbers object.) The underlying unstructured
object of an internally injective module is internally injective.
\item Any internally injective object is flabby.
% Take care if numbering changes (referenced below).
\item Any object can be embedded into a flabby object.
\item The terminal object is flabby. The product of flabby objects is flabby.
\item Let~$I$ be an internally injective object. Let~$T$ be an arbitrary
object. Then~$[T,I]$ is a flabby object.
\item (If~$\E$ has a natural numbers object.) Let~$I$ be an internally
injective~$R$-module. Let~$T$ be an arbitrary~$R$-module. Then~$[T,I]_R$, the
subobject of the internal Hom consisting only of the linear maps, is a flabby
object.
\item Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence
of~$R$-modules in~$\E$. If~$M'$ and~$M''$ are flabby objects, so is~$M$.
\end{enumerate}
\end{scholium}

@@

The notion of strongly flabby objects is, however, not stable under base change
and therefore cannot be characterized in the internal language. A specific
example is the~$G$-set~$G$ (with the translation action), considered as an
object of the topos~$BG$ of~$G$-sets, where~$G$ is a nontrivial group.
This object is not strongly flabby, since the morphism~$\emptyset \to G$ does
not lift, but its pullback to the slice~$BG/G \simeq \Set$ is (assuming
the law of excluded middle).
\end{rem}

@@

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. It is therefore instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the effective
topos. We follow the terminology of~\cite{hyland:effective-topos}.

@@

The intuitive
reason for why Proposition~\ref{prop:flabby-effective-sets} holds is the
following. Let~$X$ be a flabby object in the effective topos. Then there is a
procedure which computes for any subterminal~$K \subseteq X$ an element~$x_K$
such that~$K \subseteq \{ x_K \}$. This element might not depend extensionally
on~$K$, but this fine point is not important for this discussion. Let
now~$f : X \to X$ be a morphism. We construct the self-referential subset~$K
\defeq \{ f(x_K) \}$; the formal proof below will indicate how this can be
done. Then~$K \subseteq \{ x_K \}$, so~$f(x_K) = x_K$.

@@

This problem does not manifest with objects~$X$ which are not effective sets.
Reflexivity realizers for these objects are themselves not very informative;
a procedure witnessing flabbiness therefore only has to turn one kind of
non-informative realizers into another kind. The flabby modules featuring in
the proof of Proposition~\ref{prop:semienough-flabby-modules} will accordingly
not be effective sets.

@@

\item There is a way of embedding any module into a flabby module if one is
prepared to ignore set-theoretical difficulties. Namely, let~$M$ be an~$R$-module.
Inductively construct a collection~$T$ of terms by the following clauses: $0
\in T$ (where~$0$ is a formal symbol);
if~$t,s \in T$, then~$t + s \in T$; if~$t \in T$ and~$r \in R$, then~$rt \in
T$; if~$x \in M$, then~$\underline{x} \in T$; if~$K \subseteq T$ is a
subterminal, then~$\varepsilon_K \in T$. Let~$({\sim})$ be
the finest equivalence relation on~$T$ such that~$t + (s + u) \sim (t + s) +
u$, $t + s \sim s + t$, $t + 0 \sim t \sim 0 + t$, $1t \sim t$, $r(t+s) \sim rt
+ rs$, $(r+r')t \sim rt + r't$ and such
that~$\underline{0} \sim 0$, $\underline{x+y} \sim \underline{x} + \underline{y}$, $\underline{rx} \sim
r \underline{x}$, and such that~$\varepsilon_{\{t\}} \sim
t$ for all terms~$t$. Then the quotient~$T/{\sim}$ is a flabby~$R$-module
into which~$M$ embeds. However, it is not clear that~$T/{\sim}$ is a set.

\item There appears to be some tension regarding the effective topos:
Proposition~\ref{prop:semienough-flabby-modules} shows that at
least~$\neg\neg$-separated modules in the effective topos always embed into
flabby modules, assuming the law of excluded middle in the metatheory, while
Proposition~\ref{prop:flabby-effective-sets} shows that no nontrivial effective
module is flabby.
\end{enumerate}
